\documentclass[12pt]{article}

% ============================================================
% Packages
% ============================================================
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{float}
\usepackage{url}

\doublespacing

% ============================================================
% Title
% ============================================================
\title{Weather Shocks, Supply Dynamics, and Price Persistence: \\ Evidence from U.S.\ Lettuce Markets}
\author{Yuhan Wang\thanks{Department of Agricultural and Resource Economics, University of California, Davis. Email: yhnwang@ucdavis.edu.}}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This paper constructs a weekly dataset that integrates USDA shipment data, FOB prices, and weather data to study weather-driven supply and price dynamics in U.S.\ lettuce markets from 2010 to 2026. Using an empirical strategy that combines market-level aggregation for prediction with district-level panel variation for identification, we document three results. First, simple linear models outperform tree-based machine learning methods. OLS achieves substantially lower prediction error than XGBoost and Random Forest (RMSE \$4.36 versus \$6.67 and \$6.81), reflecting strong price persistence and a data-generating process that is largely linear conditional on lagged prices. Second, weather effects on prices are present but attenuated in aggregate data. In district-level panels, frost events increase weekly prices by \$2.33 ($p < 0.001$), while these effects are diluted in market-level regressions due to price persistence and spatial aggregation. Third, these patterns are robust across lettuce varieties, including Romaine, and across alternative validation strategies. These results show that downstream market outcomes, shipment volumes and FOB prices, can be used to identify weather-driven supply responses even in the absence of direct yield measures.
\end{abstract}

\newpage
\tableofcontents
\newpage

% ============================================================
\section{Introduction}
\label{sec:intro}
% ============================================================

Fresh produce supply chains are exposed to weather risk, but translating weather shocks into quantitative supply and price signals remains difficult. For most fresh vegetables, the data infrastructure that enables high-frequency crop monitoring in grain markets simply does not exist. There are no futures contracts, no weekly crop progress reports, and no county-level yield estimates. Buyers, distributors, and supply chain managers rely instead on trade reports, informal market intelligence, and experience. However, it remains unclear whether downstream market data can be used to recover supply responses to weather shocks.

This paper studies how weather shocks affect supply and prices in the U.S.\ iceberg lettuce market. Lettuce is the largest fresh vegetable crop in the United States by volume and is produced year-round through a seasonal rotation between California (spring through fall) and Arizona (winter). We construct a weekly dataset that integrates USDA truck shipment volumes, FOB shipping point prices, and gridded weather data from PRISM, covering six production districts over the period 2010--2026.

Our empirical strategy combines market-level aggregation for prediction with district-level panel variation for identification. At the market level, we aggregate all districts into a single weekly time series and estimate price and volume prediction models using OLS, XGBoost, and Random Forest. At the district level, we retain a three-district panel and use fixed effects to identify the causal effect of local weather shocks on prices. Both layers are necessary because aggregation improves data coverage (from 78\% to 98\% of weeks with observed prices) but dilutes the weather signal.

First, simple linear models outperform tree-based machine learning methods across all prediction tasks, including price levels, price changes, and shipment volume. The best market-level price model achieves an out-of-sample RMSE of \$4.36, a 15\% improvement over a naive autoregressive baseline. XGBoost and Random Forest perform worse than the baseline in level models and offer no improvement in change models. This reflects a data-generating process dominated by strong price persistence and approximate linearity, under which additional model complexity does not improve predictive performance and reduces robustness in extreme periods.

Second, weather effects on prices are present but obscured by price persistence in aggregate data. In level models, the three price lags together account for nearly half of total feature importance, leaving weather variables with little apparent explanatory power. When we model week-to-week price changes instead, removing the autoregressive component by construction, weather variables become substantially more important. At the district level, a frost event increases the local weekly price by \$2.33 ($p < 0.001$), while this effect is attenuated in market-level aggregates. The mechanism runs from weather to local supply disruption to local price increases, which then persist through the autoregressive structure of the market.

Third, all results are robust across lettuce varieties. Replicating the analysis with Romaine lettuce produces consistent coefficient signs and comparable, and in fact lower, prediction errors (RMSE of \$2.61 vs.\ \$4.36 for Iceberg).

The rest of the paper is organized as follows. Section~\ref{sec:literature} reviews the related literature. Section~\ref{sec:background} describes the institutional structure of the U.S.\ lettuce market. Section~\ref{sec:data} presents the data sources and variable construction. Section~\ref{sec:methods} lays out the empirical strategy. Section~\ref{sec:results} reports the main results. Section~\ref{sec:mechanism} discusses the mechanism linking weather, supply, and prices. Section~\ref{sec:robustness} presents robustness checks, and Section~\ref{sec:conclusion} concludes.

% ============================================================
\section{Literature Review}
\label{sec:literature}
% ============================================================

This paper connects four strands of literature: the econometrics of weather and crop yields, the transmission of weather shocks to agricultural prices, climate impacts on specialty crops, and the use of machine learning in applied economics.

A large literature estimates the effect of weather on agricultural output. \citet{schlenker2009nonlinear} show that the relationship between temperature and yields of corn, soybeans, and cotton in the United States is highly nonlinear: yields increase with temperature up to a crop-specific threshold and decline sharply above it. This finding established that extreme heat, rather than average warming, is the primary channel through which climate affects crop production. Subsequent work has extended this result to global scales. \citet{lobell2011climate} estimate that climate trends since 1980 have reduced global maize production by 3.8\% and wheat production by 5.5\%, while \citet{ortiz2021anthropogenic} find that anthropogenic climate change has slowed global agricultural total factor productivity growth by roughly 21\%. These studies rely on county-level or country-level yield data, typically at annual frequency. This data infrastructure does not exist for most fresh vegetables. There are no county-level yield estimates for lettuce, no weekly crop progress reports, and no futures contracts that would allow market-based inference about expected production. Our paper addresses this gap by using downstream market outcomes, shipment volumes and FOB prices, as proxies for realized supply.

A smaller literature studies how weather shocks transmit to commodity prices. \citet{roll1984orange} provides the earliest evidence, showing that frozen concentrated orange juice futures prices respond to Florida weather and even contain information beyond official weather forecasts. \citet{roberts2009world} aggregate the four major food commodities (corn, rice, soybeans, wheat) into total caloric supply and estimate supply and demand elasticities using weather-driven yield shocks as instruments. They find that demand is highly inelastic, so that modest supply disruptions produce large price responses. \citet{roberts2013identifying} formalize this identification strategy and estimate that a 1\% reduction in caloric production raises prices by roughly 2--3\%. More recently, \citet{letta2022weather} show that weather shocks affect food prices not only through realized production losses but also through shifts in trader expectations, so that prices respond to weather before the harvest occurs. Our paper complements this work by documenting a different form of attenuation: at the market level, price persistence absorbs much of the weather signal, so that the effect is visible only in price changes or in disaggregated district-level data.

Most of the weather-agriculture literature focuses on staple grains. Relatively little work has studied perishable specialty crops, despite their economic importance and their greater sensitivity to short-run weather variation. \citet{lobell2007california} is one exception: they estimate temperature and precipitation effects on yields of twelve California crops, including lettuce, using annual data from 1980 to 2003. They find that simple climate models explain over two-thirds of yield variance for most crops. \citet{kistner2018vulnerability} document that specialty crops in the Midwestern United States are more vulnerable to climatic variability than row crops, with excessive moisture, freeze events, and drought causing the largest losses. Our paper extends this work in two directions. First, we use weekly data rather than annual data, which allows us to study the short-run dynamics of weather shocks rather than long-run climate trends. Second, we work with market outcomes (prices and shipment volumes) rather than yields, which are not available at high frequency for lettuce.

Finally, there is growing interest in when machine learning methods improve on traditional econometric approaches. \citet{mullainathan2017machine} distinguish between prediction tasks, where machine learning excels, and causal inference tasks, where economic structure remains essential. \citet{storm2020machine} review applications in agricultural and applied economics, noting that machine learning is most useful when the underlying relationships are nonlinear and high-dimensional. In crop yield prediction, \citet{crane2018machine} shows that a semiparametric neural network outperforms both classical statistical methods and fully nonparametric alternatives, consistent with the nonlinear temperature-yield relationships documented by \citet{schlenker2009nonlinear}. Our results provide a counterpoint. In the lettuce price setting, tree-based methods (XGBoost and Random Forest) perform worse than OLS, both in absolute terms and relative to a naive baseline. The reason is that price dynamics are dominated by a strong autoregressive component that is approximately linear. Once this persistence is removed by differencing, all methods converge to similar performance, confirming that the residual weather-price relationship is also approximately linear. This finding contributes to the empirical record on when additional model flexibility helps and when it does not.

% ============================================================
\section{Market Background}
\label{sec:background}
% ============================================================

The U.S.\ lettuce market is highly geographically concentrated, with California and Arizona accounting for over 95\% of domestic production \citep{usda_nass_vegetables}. Production is organized around a seasonal geographic rotation between two primary regions.

\textbf{Salinas-Watsonville, California} is the dominant production district. Located in the Salinas Valley on the central California coast, it benefits from a mild maritime climate well suited to lettuce during the spring, summer, and fall months (roughly April through November, corresponding to weeks 15--50 of the year). It supplies the majority of national shipments during this period.

\textbf{Western Arizona (Yuma)} takes over during the winter months (roughly November through March, weeks 45--15). The region's warm desert climate allows lettuce production when Salinas is too cold, and shipment volumes are comparable to Salinas during peak winter weeks.

\textbf{Imperial Valley, California} serves as a supplementary desert production district, with a seasonal pattern similar to Yuma but smaller volumes. Three other California districts, Santa Maria, Oxnard, and the San Joaquin Valley, contribute smaller and more intermittent shipments.

The annual rotation between California and Arizona is largely calendar-driven and follows a stable pattern over time. Production transitions occur around weeks 15--20 (as Salinas ramps up and Arizona winds down) and weeks 45--50 (as Salinas winds down and Arizona ramps up). These transition periods are associated with temporary supply contractions and elevated price volatility. They are also particularly sensitive to weather shocks, as supply is already constrained during the shift between regions.

Lettuce has a short growing cycle of approximately 60 to 90 days from planting to harvest. Unlike annual grain crops, production occurs on a continuous, rolling basis throughout the season, with fields replanted multiple times per year. As a result, there is no single harvest event that determines total output. Instead, supply in any given week reflects planting decisions made two to three months earlier, realized growing conditions, and the pace of harvest operations. This structure generates a natural lag between weather shocks and realized supply.

This production environment has an important implication for empirical analysis. High-frequency yield data, which are available for many grain crops at the county level, do not exist for lettuce. There is no weekly or even monthly measure of production by district. Shipment data, which record the volume of lettuce loaded onto trucks at the point of origin, provide the closest available proxy for realized supply. Because lettuce is highly perishable, with a shelf life of roughly one to two weeks, storage is limited and most production is shipped shortly after harvest. Shipment volumes therefore provide a high-frequency measure of supply that is central to our empirical strategy.

% ============================================================
\section{Data}
\label{sec:data}
% ============================================================

% ------------------------------------------------------------
\subsection{Sources}
\label{subsec:sources}
% ------------------------------------------------------------

We combine three publicly available data sources covering January 2010 through April 2026.

\textbf{Shipment volume.} Weekly truck shipment reports from the USDA Agricultural Marketing Service \citep{usda_ams_truck} record the volume of each commodity shipped by truck from each production district. These reports provide the highest-frequency measure of supply available for lettuce. Unlike grain crops, which have annual county-level yield data, lettuce is harvested continuously within each growing season and no comparable high-frequency production data exist. Shipment volume therefore serves as a proxy for realized supply: because lettuce is highly perishable and cannot be stored for more than a few days, most production is shipped within the week it is harvested.

\textbf{Prices.} Daily FOB (free on board) shipping point prices from USDA Market News \citep{usda_market_news} report the price at which lettuce is sold at the point of origin, before transportation costs. We focus on FOB prices rather than retail prices because they more directly reflect production-side conditions. Prices are reported by district, commodity, and package type; we aggregate daily observations to weekly averages.

\textbf{Weather.} Daily temperature and precipitation data come from the PRISM Climate Group at Oregon State University \citep{daly2008prism}, which provides gridded climate data at 4 km resolution for the contiguous United States. For each production district, we extract daily maximum temperature, minimum temperature, and precipitation at the district centroid and aggregate these variables to weekly measures. This construction aligns weather exposure with the timing of shipment and pricing outcomes.

% ------------------------------------------------------------
\subsection{Production Districts}
\label{subsec:districts}
% ------------------------------------------------------------

The data cover six lettuce production districts in California and Arizona: Salinas-Watsonville, Western Arizona (Yuma), Imperial Valley, Santa Maria, Oxnard, and the San Joaquin Valley. Together, these districts account for the vast majority of U.S.\ lettuce production.

District activity and reporting vary substantially over time. Salinas-Watsonville, the largest district, ships in 560 of the 849 sample weeks and reports prices in 91\% of those weeks. Western Arizona ships in 429 weeks, primarily during the winter months, with 82\% price coverage. The remaining districts have more limited and irregular reporting: Oxnard reports prices in only 2\% of its active weeks, Santa Maria in 69\%, and the San Joaquin Valley in 76\%.

This uneven coverage has two implications. First, price data are missing non-randomly across districts and time. Second, relying on any single district would substantially reduce the usable sample. These features motivate the dual-layer empirical strategy described in Section~\ref{subsec:dual_layer}, which combines market-level aggregation to maximize coverage with district-level variation for identification.

% ------------------------------------------------------------
\subsection{Variable Construction}
\label{subsec:construction}
% ------------------------------------------------------------

Shipment volume is measured in units of 10,000 pounds, following USDA refrigerated truck movement reports.

\textbf{Market-level price.} For each week, we construct a shipment-weighted average FOB price across all districts that report a price (Equation~\ref{eq:weighted_price}). Weighting by shipment volume ensures that districts with larger production have greater influence on the market price. To account for missing price observations, we define a coverage ratio (Equation~\ref{eq:coverage}) as the share of total weekly shipments for which a price is observed. The mean coverage ratio is 0.933, and 98.2\% of weeks have at least one district reporting a price. This construction yields a consistent weekly price series while explicitly tracking the reliability of the underlying price information.

\textbf{Market-level weather.} Weather variables are aggregated to the market level using shipment weights. Market-level weather reflects conditions in the districts that are actively supplying the market in a given week, rather than a fixed geographic average. When production shifts seasonally from Salinas to Yuma, the market-level weather measures shift accordingly.

\textbf{Extreme weather indicators.} Using the daily weather data, we construct three indicator variables for each district-week. The variable $freeze\_risk$ equals one if the minimum temperature falls below 0\textdegree C on any day within the week. The variable $extreme\_heat$ equals one if the maximum temperature exceeds 40\textdegree C on any day within the week. The variable $heavy\_rain$ equals one if total weekly precipitation exceeds 25 mm. At the market level, each indicator equals one if any active district experiences the event.

% ------------------------------------------------------------
\subsection{Summary Statistics}
\label{subsec:summary}
% ------------------------------------------------------------

Table~\ref{tab:summary} reports summary statistics for the market-level dataset. The mean weekly iceberg lettuce FOB price is \$16.27 per carton with a standard deviation of \$11.80. The distribution is right-skewed, with a median of \$12.30 and a maximum of \$94.33, reached during the 2022 INSV outbreak. Mean weekly shipment volume is 4,422 (in 10,000 lb units). Extreme weather events are relatively infrequent at the market level: freeze risk occurs in 14 of 849 weeks (1.6\%), extreme heat in 6 weeks (0.7\%), and heavy rain in 68 weeks (8.0\%).

\begin{table}[H]
\centering
\caption{Summary statistics: Market-level dataset (849 weekly observations, 2010--2026)}
\label{tab:summary}
\begin{tabular}{lrrrr}
\toprule
Variable & Mean & SD & Min & Max \\
\midrule
Price (\$/carton) & 16.27 & 11.80 & 5.99 & 94.33 \\
Volume (10,000 lbs) & 4,422 & 749 & 2,313 & 6,993 \\
Coverage ratio & 0.93 & 0.19 & 0.00 & 1.00 \\
Tmax (\textdegree C) & 21.90 & 3.20 & 14.80 & 35.70 \\
Tmin (\textdegree C) & 10.40 & 2.50 & 1.90 & 17.70 \\
Precipitation (mm) & 1.95 & 5.93 & 0.00 & 80.52 \\
Active districts & 2.90 & 1.10 & 2.00 & 6.00 \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Methodology and Empirical Strategy}
\label{sec:methods}
% ============================================================

This section describes the empirical framework. Section~\ref{subsec:dual_layer} outlines the dual-layer design that motivates the two datasets. Sections~\ref{subsec:market_model}--\ref{subsec:district_panel} present the model specifications. Section~\ref{subsec:features} describes feature construction, and Section~\ref{subsec:validation} details the out-of-sample validation strategy.

% ------------------------------------------------------------
\subsection{Dual-Layer Design}
\label{subsec:dual_layer}
% ------------------------------------------------------------

The U.S.\ lettuce market spans six production districts across California and Arizona, but not all districts report FOB prices every week. In particular, the three smaller districts (Santa Maria, Oxnard, and San Joaquin Valley) exhibit substantial gaps in price reporting. A district-level panel including all six districts would have a price coverage rate of only 78.4\%, making it unsuitable for time-series prediction.

We address this by constructing two datasets from the same underlying weekly panel. The first aggregates all six districts into a single market-level time series. The second retains the three core districts (Salinas-Watsonville, Western Arizona, and Imperial Valley) as a panel.

\textbf{Dataset A: Market-level aggregate.} Each week, we compute a shipment-weighted average price across all districts that report a price:
\begin{equation}
\label{eq:weighted_price}
P_t = \frac{\sum_{i} V_{it} \cdot P_{it}}{\sum_{i} V_{it}},
\end{equation}
where $V_{it}$ is iceberg lettuce shipment volume from district $i$ in week $t$, and the sum runs over districts with nonmissing prices. Weather variables are aggregated analogously using shipment weights. This produces 849 weekly observations with 98.2\% price coverage. The coverage ratio,
\begin{equation}
\label{eq:coverage}
C_t = \frac{\sum_{i \in \text{priced}} V_{it}}{\sum_{i} V_{it}},
\end{equation}
measures the share of total shipments for which a price is observed and serves as a data quality indicator.

\textbf{Dataset B: District-level panel.} We retain the three districts with both substantial shipment volume and relatively consistent price reporting: Salinas-Watsonville, Western Arizona, and Imperial Valley. This panel contains 1,466 observations with 78.4\% price coverage.

The two datasets serve distinct roles. Dataset A is used for prediction, where complete coverage and a continuous time series are essential. Dataset B is used for identification, where cross-district variation in weather provides the identifying variation.

% ------------------------------------------------------------
\subsection{Market-Level Prediction Model}
\label{subsec:market_model}
% ------------------------------------------------------------

The market-level price model takes the form
\begin{equation}
\label{eq:market_level}
P_t = \alpha + \boldsymbol{\beta}_1' \mathbf{W}_t + \boldsymbol{\beta}_2' \mathbf{L}_t + \boldsymbol{\beta}_3' \mathbf{S}_t + \boldsymbol{\beta}_4' \mathbf{X}_t + \varepsilon_t,
\end{equation}
where $P_t$ is the shipment-weighted FOB price in week $t$, $\mathbf{W}_t$ is a vector of contemporaneous weather variables, $\mathbf{L}_t$ contains lagged price and volume terms, $\mathbf{S}_t$ includes seasonal indicators, and $\mathbf{X}_t$ contains controls such as the coverage ratio and the number of active districts.

We estimate Equation~\eqref{eq:market_level} by OLS with heteroskedasticity-robust standard errors \citep{white1980heteroskedasticity}. We also estimate the same specification using XGBoost \citep{chen2016xgboost} and Random Forest \citep{breiman2001random} as nonlinear benchmarks.

For the tree-based models, we use conservative hyperparameters to limit overfitting. The XGBoost model uses 200 trees, maximum depth 3, learning rate 0.03, subsample ratio 0.7, and L1/L2 regularization ($\alpha = 1.0$, $\lambda = 5.0$). The Random Forest uses 500 trees, maximum depth 6, and minimum leaf size 10. These settings are intentionally restrictive, as the goal is to evaluate out-of-sample performance rather than maximize in-sample fit.

% ------------------------------------------------------------
\subsection{Change Model}
\label{subsec:change_model}
% ------------------------------------------------------------

Lettuce prices are highly persistent: the first-order autocorrelation exceeds 0.9, and in level regressions the lagged price explains most of the variation. This persistence can mask the effects of other variables, including weather. To isolate the role of short-run shocks, we estimate models in first differences:
\begin{equation}
\label{eq:change_model}
\Delta P_t = P_t - P_{t-1} = \alpha + \boldsymbol{\gamma}' \mathbf{W}_t + \boldsymbol{\delta}' \mathbf{M}_t + u_t,
\end{equation}
where $\mathbf{W}_t$ contains weather variables and $\mathbf{M}_t$ includes volume, coverage, and seasonal controls. Lagged prices are excluded because differencing removes the autoregressive component.

Comparing the level and change specifications allows us to assess whether weather effects are attenuated by price persistence.

% ------------------------------------------------------------
\subsection{District-Level Panel}
\label{subsec:district_panel}
% ------------------------------------------------------------

The district panel exploits cross-sectional variation in weather exposure that is averaged out at the market level. A freeze event in Yuma, for example, affects local prices directly but is diluted when aggregated with regions that do not experience the same shock.

Identification relies on the assumption that short-run weather variation affects supply but not demand. Weather shocks influence prices through crop damage and harvest disruptions rather than contemporaneous shifts in consumer demand. This assumption is standard in the agricultural economics literature and is plausible at the weekly frequency considered here.

The district panel specification is
\begin{equation}
\label{eq:district_change}
\Delta P_{it} = \boldsymbol{\phi}' \mathbf{W}_{it} + \alpha_i + \mu_q + \eta_{iq} + \nu_{it},
\end{equation}
where $\Delta P_{it}$ is the week-over-week price change for district $i$ in week $t$, $\mathbf{W}_{it}$ is the local weather vector, $\alpha_i$ are district fixed effects, $\mu_q$ are quarter fixed effects, and $\eta_{iq}$ captures district-by-season interactions. Standard errors are clustered at the district level. Identification comes from cross-district variation in weather within the same week.

We also estimate a level specification that includes lagged prices:
\begin{equation}
\label{eq:district_level}
P_{it} = \boldsymbol{\phi}' \mathbf{W}_{it} + \boldsymbol{\psi}' \mathbf{L}_{it} + \alpha_i + \mu_q + \eta_{iq} + \nu_{it}.
\end{equation}
Comparing the level and change specifications at the district level provides an additional check on whether persistence attenuates estimated weather effects.

% ------------------------------------------------------------
\subsection{Feature Construction}
\label{subsec:features}
% ------------------------------------------------------------

All models use the same set of constructed features.

\textbf{Weather variables.} Weekly weather includes maximum and minimum temperature ($tmax$, $tmin$), total precipitation ($ppt$), and diurnal range ($tmax - tmin$). Three indicators capture extreme conditions: $freeze\_risk$ equals one if minimum temperature falls below 0\textdegree C during the week, $extreme\_heat$ equals one if maximum temperature exceeds 40\textdegree C, and $heavy\_rain$ equals one if weekly precipitation exceeds 25 mm.

\textbf{Lagged variables.} For prices, shipment volume, temperature, and precipitation, we include lags at 1, 2, and 4 weeks to capture short-run persistence and the production cycle.

\textbf{Rolling averages.} Four-week rolling means of volume and weather variables capture the recent background state of the market.

\textbf{Market structure.} The coverage ratio $C_t$ and the number of active districts $n_t$ enter as controls.

\textbf{Seasonality.} Month and week-of-year enter as continuous variables to capture the strong seasonal pattern in production and prices.

% ------------------------------------------------------------
\subsection{Validation Strategy}
\label{subsec:validation}
% ------------------------------------------------------------

Out-of-sample performance is evaluated using an expanding-window approach. For each test year $\tau$ from 2014 to 2026, models are trained on all prior years and evaluated on year $\tau$, mimicking a real-time forecasting setting.

We require a minimum of four years of training data, so the first test year is 2014 (trained on 2010--2013). The expanding window is preferred over a rolling window because the sample is relatively short and additional training data improves stability; we verify this in the diagnostics section.

Model performance is measured using root mean squared error (RMSE) in price levels. For change models, predictions are converted back to price levels before evaluation. The naive baseline predicts $\hat{P}_t = P_{t-1}$, corresponding to zero predicted price change.

% ============================================================
\section{Results}
\label{sec:results}
% ============================================================

% ------------------------------------------------------------
\subsection{Market-Level Price Prediction}
\label{subsec:price_prediction}
% ------------------------------------------------------------

Table~\ref{tab:model_comparison} reports average out-of-sample prediction errors for the market-level price models over the 2014--2026 test period. The naive baseline, which predicts $\hat{P}_t = P_{t-1}$, has an average RMSE of \$5.11. The OLS level model with the full feature set achieves an RMSE of \$4.36, a 14.7\% improvement over the naive baseline and the best performance among the models considered.

Both tree-based methods perform worse than OLS in level models. XGBoost averages \$6.67 and Random Forest \$6.81, each substantially worse than the naive baseline. This pattern is not driven by hyperparameter choice. The conservative settings we use are designed to limit overfitting, and the models capture meaningful in-sample variation. Rather, lettuce prices are driven primarily by a strong autoregressive component, and tree-based models do not perform well in this setting out of sample.

The year-by-year results reveal the source of the gap most clearly in 2022. That year, an outbreak of Impatiens Necrotic Spot Virus (INSV) in the Salinas Valley disrupted production severely, and the weekly FOB price for iceberg lettuce reached \$94.33, roughly five times the sample mean. OLS tracked the spike with an RMSE of \$7.77 for that year, while XGBoost and Random Forest produced much larger errors, with RMSEs of \$24.37 and \$23.83, respectively. Because the training data contain no comparable price spike, the tree-based models substantially underpredict the 2022 peak. This illustrates a limitation of tree-based methods in small-sample time series settings: they are generally poor at extrapolating beyond the range of the training data, while linear autoregressive models can extend trends through the lagged price coefficient.

\begin{table}[H]
\centering
\caption{Out-of-sample model comparison: average RMSE and MAE (2014--2026, expanding window)}
\label{tab:model_comparison}
\begin{tabular}{lrrr}
\toprule
Model & Avg RMSE (\$) & Avg MAE (\$) & Improvement vs.\ Naive \\
\midrule
Naive ($P_t = P_{t-1}$) & 5.11 & 3.35 & baseline \\
\addlinespace
\textit{Level models} \\
\quad OLS (full) & \textbf{4.36} & 3.14 & +14.7\% \\
\quad XGBoost & 6.67 & 4.26 & $-$30.5\% \\
\quad Random Forest & 6.81 & 4.32 & $-$33.3\% \\
\addlinespace
\textit{Change models} \\
\quad OLS $\Delta P$ & 5.07 & 3.57 & +0.8\% \\
\quad XGBoost $\Delta P$ & 5.12 & 3.56 & $-$0.2\% \\
\quad Random Forest $\Delta P$ & 5.00 & 3.40 & +2.2\% \\
\quad OLS log $\Delta P$ & 5.00 & 3.41 & +2.2\% \\
\bottomrule
\end{tabular}
\end{table}

% ------------------------------------------------------------
\subsection{Price Persistence and Hidden Weather Effects}
\label{subsec:persistence}
% ------------------------------------------------------------

In the full level regression, the lagged price coefficient is 1.365 ($p < 0.001$), and the in-sample $R^2$ is 0.891. The one-week lagged price alone accounts for 28\% of total feature importance in the XGBoost level model, and the three price lags together account for 47\%, indicating that price persistence dominates the predictive structure. Precipitation is statistically significant ($\hat{\beta} = 0.123$, $p = 0.010$), but its economic magnitude is small relative to the lagged price term.

The change models tell a different story. When we difference out price persistence by modeling $\Delta P_t$ instead of $P_t$, two things happen. First, the three estimation methods converge: OLS, XGBoost, and Random Forest all produce average RMSEs within a narrow range of \$5.00 to \$5.13. The gap between OLS and tree-based methods disappears because the dominant linear feature, lagged price, has been removed. The residual relationship between weather and price changes is approximately linear, so nonlinear methods offer no advantage.

Second, and more importantly, the relative importance of weather variables increases substantially. In the level model, temperature ($tmax$) contributes only 2.4\% of feature importance. In the change model, where price lags are excluded by construction, temperature importance rises to 7.5\%, and other weather variables (precipitation, diurnal range) increase as well. The weather signal was present in the data but was masked by price persistence in the level specification.

% ------------------------------------------------------------
\subsection{Volume Prediction}
\label{subsec:volume}
% ------------------------------------------------------------

Table~\ref{tab:volume} reports results for market-level shipment volume prediction. The pattern mirrors the price results: OLS outperforms XGBoost. The OLS level model achieves an average RMSE of 287, a 38.0\% improvement over the naive baseline of 463. XGBoost achieves 354, better than the naive baseline but well behind OLS.

Precipitation is the most consistent weather predictor of volume. In the OLS level model, the coefficient on weekly precipitation is $-9.0$ (in 10,000 lb units) per millimeter ($p = 0.015$). Heavy rain disrupts field operations and harvest logistics, reducing the volume of lettuce that can be shipped in a given week. Freeze risk has a large negative point estimate ($-195$) but is not statistically significant at conventional levels ($p = 0.12$), likely because freeze events are infrequent in the sample.

\begin{table}[H]
\centering
\caption{Volume prediction: average RMSE and MAE (10,000 lbs, expanding window, 2014--2026)}
\label{tab:volume}
\begin{tabular}{lrrr}
\toprule
Model & Avg RMSE (10,000 lbs) & Avg MAE (10,000 lbs) & Improvement vs.\ Naive \\
\midrule
Naive ($V_t = V_{t-1}$) & 463 & 347 & baseline \\
OLS Level & \textbf{287} & 223 & +38.0\% \\
XGBoost Level & 354 & 265 & +23.6\% \\
OLS $\Delta V$ & 459 & 345 & +0.8\% \\
XGBoost $\Delta V$ & 453 & 345 & +2.1\% \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Mechanism: Weather, Supply, and Price}
\label{sec:mechanism}
% ============================================================

The results in Section~\ref{sec:results} establish that weather matters for lettuce prices but that its effect is masked by price persistence in level models. This section uses the district-level panel to identify how weather operates, and to explain why the market-level aggregate understates weather effects.

% ------------------------------------------------------------
\subsection{District-Level Identification}
\label{subsec:district_id}
% ------------------------------------------------------------

Table~\ref{tab:district_weather} reports weather coefficients from the district panel specifications described in Section~\ref{subsec:district_panel}. The change model (Equation~\ref{eq:district_change}) includes district fixed effects, quarter fixed effects, and district-by-season interactions. Standard errors are clustered at the district level.

Freeze risk is the strongest and most robust weather predictor. In the change model, a week with at least one day below freezing is associated with a price increase of \$2.33 ($p < 0.001$). This estimate is stable across specifications: adding volume controls and lagged weather changes the coefficient only slightly, from \$2.33 to \$2.37 to \$2.24. Minimum temperature has a significant negative coefficient ($-\$0.15$ to $-\$0.17$ per degree Celsius, $p < 0.01$), which is the continuous counterpart of the freeze risk indicator: colder weeks are associated with higher prices. Extreme heat is significant in the level specification (\$2.16, $p = 0.008$) but not in the change models, suggesting that heat effects operate over a slightly longer horizon than week-to-week changes can capture.

\begin{table}[H]
\centering
\caption{District-level weather coefficients (Iceberg, OLS with clustered SE)}
\label{tab:district_weather}
\begin{tabular}{lrrrr}
\toprule
Variable & $\Delta P_{it}$ (M1) & $\Delta P_{it}$ + vol (M2) & + weather lags (M3) & Level (M4) \\
\midrule
freeze\_risk & 2.326$^{***}$ & 2.368$^{***}$ & 2.237$^{***}$ & 1.354 \\
tmin\_avg & $-$0.164$^{**}$ & $-$0.169$^{***}$ & $-$0.145$^{***}$ & $-$0.128$^{***}$ \\
extreme\_heat & 0.542 & 0.493 & 0.778 & 2.164$^{***}$ \\
ppt\_total & 0.106 & 0.105 & 0.104 & 0.089 \\
heavy\_rain & $-$0.914 & $-$0.919 & $-$0.824 & $-$2.124 \\
\bottomrule
\multicolumn{5}{l}{\footnotesize $^{***}p<0.01$, $^{**}p<0.05$, $^{*}p<0.10$. All models include district FE, quarter FE,} \\
\multicolumn{5}{l}{\footnotesize and district $\times$ season interactions. SE clustered by district.} \\
\end{tabular}
\end{table}

The precipitation coefficient is positive but not statistically significant in any specification. This result is somewhat unexpected given that precipitation significantly predicts both price at the market level and volume. A plausible explanation is that precipitation effects on price operate primarily through volume reductions at the market level, and the district panel, which does not aggregate across districts, captures a noisier version of this channel.

% ------------------------------------------------------------
\subsection{Why Level Models Mask Weather Effects}
\label{subsec:why_masked}
% ------------------------------------------------------------

The comparison across columns in Table~\ref{tab:district_weather} illustrates how price persistence attenuates weather coefficients. Freeze risk is large and highly significant in all three change specifications but shrinks and loses significance in the level model (column M4). The level model includes lagged prices, which absorb much of the variation that weather explains in the change model. Since a freeze event raises prices this week, and lagged price carries that increase forward into next week, the level model attributes the persistent effect to the lag rather than to the freeze.

This attenuation is not a statistical artifact. It reflects a real feature of the price process. Weather shocks are transitory, but their price effects persist through the autoregressive structure of the market. A freeze in Yuma raises the Yuma price, which feeds into the market-level price, which then carries forward through price persistence. The level model captures the persistence but not the initial cause. The change model captures the cause but not the persistence. Both are needed to tell the full story.

The district panel also reveals heterogeneity that is lost in the market aggregate. A freeze event in Western Arizona has a direct and immediate effect on the local price, but when averaged with Salinas (which may not experience a freeze in the same week), the signal is diluted. This helps explain why freeze risk is highly significant ($p < 0.001$) in the district panel but not significant in the market-level regressions. The dual-layer design is not just a convenience for handling missing data; it is necessary for identifying localized weather effects that aggregation would obscure.

Taken together, the evidence supports the following mechanism. Weather shocks, particularly freezes and heat waves, reduce local supply in the affected district. The supply reduction raises the local FOB price. Because the market is spatially connected (buyers can substitute across districts to some extent), the price increase propagates to the market level. Once embedded in the market price, the shock persists through the autoregressive price process. The chain runs: weather $\to$ local supply $\to$ local price $\to$ market price $\to$ persistence. This mechanism reconciles the apparent disconnect between weak weather effects in aggregate regressions and strong effects in the panel estimates.

% ============================================================
\section{Robustness}
\label{sec:robustness}
% ============================================================

% ------------------------------------------------------------
\subsection{Romaine Lettuce}
\label{subsec:romaine}
% ------------------------------------------------------------

We replicate the core analysis using Romaine lettuce prices instead of Iceberg. Table~\ref{tab:romaine} compares the two varieties. The market-level OLS model achieves an out-of-sample RMSE of \$2.61 for Romaine, compared to \$4.36 for Iceberg. The lower RMSE reflects lower price volatility rather than improved model fit; the in-sample $R^2$ is actually higher for Romaine (0.952 vs.\ 0.891).

\begin{table}[H]
\centering
\caption{Romaine vs.\ Iceberg: model performance and weather coefficients}
\label{tab:romaine}
\begin{tabular}{lrr}
\toprule
 & Iceberg & Romaine \\
\midrule
\textit{Model performance} \\
\quad In-sample $R^2$ (level) & 0.891 & 0.952 \\
\quad In-sample RMSE & \$3.97 & \$2.48 \\
\quad OOS avg RMSE & \$4.36 & \$2.61 \\
\quad Price mean & \$16.27 & \$17.80 \\
\quad Price coverage & 98.2\% & 98.4\% \\
\addlinespace
\textit{Key coefficients} \\
\quad price\_lag1 (level) & 1.369$^{***}$ & 1.633$^{***}$ \\
\quad ppt\_total ($\Delta P$) & 0.192$^{***}$ & 0.137$^{***}$ \\
\quad freeze\_risk ($\Delta P$) & 1.176 & 1.843$^{**}$ \\
\quad heavy\_rain ($\Delta P$) & $-$2.003$^{**}$ & $-$1.647$^{***}$ \\
\bottomrule
\multicolumn{3}{l}{\footnotesize $^{***}p<0.01$, $^{**}p<0.05$, $^{*}p<0.10$.} \\
\end{tabular}
\end{table}

More importantly, the weather coefficients have consistent signs across the two varieties. Precipitation is positive and significant for both ($p < 0.01$), freeze risk is positive in both (and significant for Romaine), and heavy rain is negative and significant in both. The core findings of this paper, that weather affects prices through supply disruptions and that these effects are masked by price persistence, are not specific to Iceberg lettuce. They reflect the structure of the lettuce market more broadly.

% ------------------------------------------------------------
\subsection{Expanding vs.\ Rolling Window}
\label{subsec:window}
% ------------------------------------------------------------

Table~\ref{tab:window} compares expanding-window and five-year rolling-window validation. The expanding window produces a lower average RMSE (\$4.36 vs.\ \$4.57). The two approaches yield nearly identical results in early test years, when they share the same training set. The gap widens in later years: in 2021, the expanding window achieves \$3.41 vs.\ \$4.61 for rolling, and in 2022 the gap is \$7.77 vs.\ \$9.22. The expanding window benefits from a larger training sample, and the data-generating process appears sufficiently stable that older observations remain informative.

\begin{table}[H]
\centering
\caption{Expanding vs.\ rolling window: year-by-year RMSE (\$)}
\label{tab:window}
\begin{tabular}{lrrrr}
\toprule
 & \multicolumn{2}{c}{Expanding} & \multicolumn{2}{c}{Rolling (5yr)} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
Year & RMSE & MAE & RMSE & MAE \\
\midrule
2014 & 2.47 & 1.86 & 2.47 & 1.86 \\
2015 & 3.60 & 2.83 & 3.60 & 2.83 \\
2016 & 2.66 & 1.84 & 2.71 & 1.85 \\
2017 & 4.92 & 3.40 & 5.00 & 3.48 \\
2018 & 5.12 & 2.94 & 5.00 & 2.88 \\
2019 & 4.61 & 3.71 & 4.69 & 3.78 \\
2020 & 4.17 & 2.95 & 4.27 & 2.97 \\
2021 & 3.41 & 2.71 & 4.61 & 3.92 \\
2022 & 7.77 & 4.93 & 9.22 & 5.81 \\
2023 & 4.62 & 3.16 & 4.67 & 3.15 \\
2024 & 4.56 & 3.53 & 4.71 & 3.54 \\
2025 & 4.16 & 2.97 & 4.12 & 2.88 \\
2026 & 4.59 & 4.04 & 4.28 & 3.77 \\
\midrule
\textbf{Average} & \textbf{4.36} & \textbf{3.14} & \textbf{4.57} & \textbf{3.29} \\
\bottomrule
\end{tabular}
\end{table}

% ------------------------------------------------------------
\subsection{Coverage and Prediction Error}
\label{subsec:coverage_error}
% ------------------------------------------------------------

Table~\ref{tab:coverage} shows that weeks with low price coverage are harder to predict. When the coverage ratio falls below 0.5, the average MAE is \$4.98, nearly double the \$2.83 for weeks with coverage above 0.95. This pattern is expected: when fewer districts report prices, the shipment-weighted average is noisier, and the model has less reliable training signal. Including the coverage ratio as a feature helps the model partially adjust for this, but it cannot fully compensate for missing price information.

The low-coverage bin (0.5--0.7) contains only one observation and should be interpreted with caution.

\begin{table}[H]
\centering
\caption{Prediction error by coverage ratio}
\label{tab:coverage}
\begin{tabular}{lrrrr}
\toprule
Coverage bin & $N$ & MAE (\$) & Median error (\$) & MAPE (\%) \\
\midrule
Very low ($< 0.5$) & 17 & 4.98 & 4.43 & 29.6 \\
Low (0.5--0.7) & 1 & 3.97 & 3.97 & 16.8 \\
Medium (0.7--0.85) & 69 & 3.55 & 2.56 & 18.2 \\
High (0.85--0.95) & 69 & 3.99 & 2.87 & 19.5 \\
Very high ($> 0.95$) & 480 & 2.83 & 1.84 & 18.1 \\
\bottomrule
\end{tabular}
\end{table}

% ------------------------------------------------------------
\subsection{The 2022 Extreme Year}
\label{subsec:extreme_2022}
% ------------------------------------------------------------

The 2022 crop year was an outlier by any measure. Table~\ref{tab:extreme2022} summarizes the key statistics. Weekly iceberg lettuce prices averaged \$31.89 with a standard deviation of \$27.09, compared to typical years where the mean is around \$15 and the standard deviation is around \$5. The maximum weekly price reached \$94.33, driven by an outbreak of Impatiens Necrotic Spot Virus (INSV) in the Salinas Valley that severely disrupted production.

\begin{table}[H]
\centering
\caption{2022 extreme year: model performance comparison}
\label{tab:extreme2022}
\begin{tabular}{lrr}
\toprule
 & 2022 & Other years \\
\midrule
\textit{Price statistics} \\
\quad Mean price & \$31.89 & $\sim$\$15 \\
\quad Max price & \$94.33 & $\sim$\$50 \\
\quad Price std & \$27.09 & $\sim$\$5 \\
\addlinespace
\textit{Model RMSE} \\
\quad OLS Level & \$7.77 & \$3.95 \\
\quad XGBoost Level & \$24.37 & \$4.75 \\
\quad Random Forest Level & \$23.83 & \$4.89 \\
\addlinespace
\textit{Prediction error (OLS)} \\
\quad MAE & \$4.93 & \$2.93 \\
\bottomrule
\end{tabular}
\end{table}

The disparity between OLS and tree-based methods in 2022 highlights an extrapolation issue: tree-based models are generally unable to extrapolate beyond the range of their training data, while OLS can follow the upward movement through the autoregressive component. We do not interpret 2022 as a failure of the model. Extreme phytosanitary shocks are inherently difficult to forecast using weather and market data alone. The relevant finding is that OLS degrades more gracefully under extreme conditions, rather than producing highly unstable predictions.

% ------------------------------------------------------------
\subsection{Feature Stability}
\label{subsec:stability}
% ------------------------------------------------------------

We split the sample into three sub-periods and re-estimate the market-level OLS model on each. Table~\ref{tab:stability} reports the key coefficients. The lagged price coefficient ranges from 1.153 to 1.517 across periods, and the precipitation coefficient from 0.116 to 0.181. There is no clear evidence of a structural break in the core relationships.

The temperature variables ($tmax$, $tmin$) show sign instability, which likely reflects collinearity between them rather than instability in the underlying relationship. Freeze risk also varies across periods, consistent with its low frequency making it sensitive to individual events within each sub-sample.

\begin{table}[H]
\centering
\caption{Coefficient stability across sub-periods (market-level OLS)}
\label{tab:stability}
\begin{tabular}{lrrr}
\toprule
Variable & 2010--2015 & 2016--2020 & 2021--2026 \\
\midrule
price\_lag1 & 1.218 & 1.153 & 1.517 \\
ppt\_total & 0.116 & 0.118 & 0.181 \\
freeze\_risk & 1.682 & $-$1.793 & $-$1.355 \\
heavy\_rain & $-$2.190 & $-$0.670 & $-$2.181 \\
extreme\_heat & $-$0.000 & $-$0.000 & $-$0.160 \\
\bottomrule
\end{tabular}
\end{table}

% ------------------------------------------------------------
\subsection{Missing Price Prediction}
\label{subsec:missing_price}
% ------------------------------------------------------------

Not all district-week observations have reported prices. Out of 2,441 district-week observations with positive shipment volume, 682 (28\%) are missing a price. We train a model on observed prices using weather, volume, district indicators, and seasonal controls, then predict prices for the missing observations. OLS achieves an average out-of-sample RMSE of \$11.08 (expanding-window validation), compared to \$13.27 for XGBoost.

Table~\ref{tab:missing} compares predicted and observed mean prices by district. The predicted prices for missing observations are systematically higher than the observed means. For example, the predicted mean for Oxnard's missing weeks is \$17.72 vs.\ an observed mean of \$15.30. This pattern is consistent with non-random missingness in price reporting: districts may be less likely to report prices during periods of high prices or supply stress.

\begin{table}[H]
\centering
\caption{Missing price prediction by district}
\label{tab:missing}
\begin{tabular}{lrrr}
\toprule
District & Missing obs & Predicted mean (\$) & Observed mean (\$) \\
\midrule
Oxnard & 238 & 17.72 & 15.30 \\
Santa Maria & 199 & 19.57 & 13.86 \\
Western Arizona & 76 & 18.93 & 17.26 \\
Imperial Valley & 71 & 20.93 & 15.49 \\
Salinas-Watsonville & 49 & 23.72 & 16.14 \\
San Joaquin Valley & 49 & 22.18 & 20.40 \\
\midrule
\textbf{Total} & \textbf{682} & & \\
\bottomrule
\multicolumn{4}{l}{\footnotesize OLS RMSE = \$11.08, XGBoost RMSE = \$13.27 (expanding window).} \\
\end{tabular}
\end{table}

% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

This paper constructs a weekly data pipeline that integrates USDA shipment data, FOB prices, and PRISM weather data to study how weather shocks affect supply and prices in U.S.\ lettuce markets.

We document three main results. First, simple linear models outperform tree-based machine learning methods for price prediction. OLS achieves an out-of-sample RMSE of \$4.36, compared to \$6.67 for XGBoost and \$6.81 for Random Forest. Lettuce prices exhibit strong persistence, and most of the predictable variation is captured by lagged prices. Conditional on this persistence, the remaining relationship between weather and prices is approximately linear, so more flexible methods do not improve predictive performance and perform poorly during extreme events.

Second, weather effects on prices are real but masked by price persistence. In level models, the three price lags together account for nearly half of total feature importance, leaving limited scope for weather variables to contribute. When we model price changes instead of price levels, weather variables become substantially more important. At the district level, freeze risk increases weekly prices by \$2.33 ($p < 0.001$), but this effect is diluted when aggregated to the market level. The dual-layer approach, combining market-level aggregation for prediction with district-level panels for identification, is therefore essential for uncovering localized weather effects.

Third, these findings are robust. The results replicate for Romaine lettuce, coefficients are stable across sub-periods, and alternative validation strategies yield similar conclusions. Diagnostics also show that data quality, as measured by price coverage, is an important determinant of prediction accuracy.

Several limitations remain. Identification relies on the assumption that short-run weather variation affects supply but not demand. While plausible at the weekly frequency considered here, longer-run climate trends may affect both sides of the market. The 2022 virus outbreak highlights that non-weather shocks can dominate market outcomes, and such events are inherently difficult to predict using historical price and weather data. In addition, the analysis focuses on six production districts in California and Arizona; extending the framework to other crops or regions would require additional data and may yield different dynamics.

Despite these limitations, the results show that downstream market outcomes, shipment volumes and FOB prices, can be used to infer climate-driven supply risk even in the absence of high-frequency yield data. For participants in fresh produce markets, the combination of publicly available USDA data and simple linear models provides a transparent and practical approach to monitoring weather-related price risk.

% ============================================================
% References
% ============================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
